---
title: "CSC_D3S4 - Spatial modelling with rasters"
author: "Marko Kallio"
date: "3 Nov 2021"
output: html_document
---



In the last session of the course, we'll do a some simple modelling with rasters.


Load temperature and rainfall stacks, elevation, and lake surfaces.

```{r}

library(terra)
library(dplyr)
library(sf)

files <- list.files("/home/rstudio/r-spatial-course/Data/raster/RRmon_Monthly_2005-2014/", 
                    full.names = TRUE)
rainfall <- lapply(files, terra::rast) #read in all files, output is a list
rainfall <- do.call("c", rainfall) # stack all rasters together

files <- list.files("/home/rstudio/r-spatial-course/Data/raster/Tmean_Monthly_2005-2014/", 
                    full.names = TRUE)

temperature <- lapply(files, terra::rast) #read in all files, output is a list
temperature <- do.call("c", temperature) # stack all rasters together

elevation <- terra::rast("/home/rstudio/r-spatial-course/Data/raster/Elevation_10km.tif") # Digital elevation model
lakes <- terra::rast("/home/rstudio/r-spatial-course/Data/raster/Lake_10km.tif") # Lake cover


```



Let's run a linear model looking at the changes in temperature and rainfall in 2005-2014. We'll define a function extracting the model coefficients.


```{r}


### Define functions to fit a model and extract relevant information ###
# For obtaining estimated model coefficients
fun_model <- function(x) { 
    
    if(all(is.na(x))) return(c(NA,NA,NA,NA))
    
    n <- 1:length(x)
    linmod <- lm(x ~ n)
    coefs <- linmod$coefficients
    pvals <- summary(linmod)$coefficients[,4]
    out <- c(coefs, pvals)
    
    return(out)
}

rainfall_coef <- terra::app(rainfall, fun = fun_model)
rainfall_coef
names(rainfall_coef) <- c("Intercept", "Slope", "p_int", "p_slope")
plot(rainfall_coef)

temperature_coef <- terra::app(temperature, fun = fun_model)
names(temperature_coef) <- c("Intercept", "Slope", "p_int", "p_slope")
plot(temperature_coef)


```


Let's look at the trend in a single month, February. 


```{r}
subset_vect <- seq(2, by = 12, length.out = 10)
subset_vect

feb_temp <- terra::subset(temperature, subset = subset_vect)
feb_rain <- terra::subset(rainfall, subset = subset_vect)


rainfall_coef <- terra::app(feb_rain, fun = fun_model)
names(rainfall_coef) <- c("Intercept", "Slope", "p_int", "p_slope")
plot(rainfall_coef)

temperature_coef <- terra::app(feb_temp, fun = fun_model)
names(temperature_coef) <- c("Intercept", "Slope", "p_int", "p_slope")
plot(temperature_coef)


```


There seems to be significant warming in most of Finland over the period 2005-2014. We could also extract the values as a data.frame.


```{r}

model_data <- as.data.frame(temperature_coef, 
                            xy = TRUE,
                            cells = TRUE) %>% 
    dplyr::as_tibble()

summary(model_data)
model_data
```




## Applying a pre-specified model to a raster stack


Let's fit a linear model predicting a average temperature in July 2011 based on location (X, Y), and elevation. We train it with some random selection of training data and assess against the remaining testing data.

```{r}
july2011 <- temperature[["Tmon_20110701"]]
july2011 <- terra::project(july2011, elevation, method = "near")

stack <- c(july2011, elevation)
plot(stack)

data <- terra::as.data.frame(stack, xy = TRUE)
data
trainset <- sample(1:nrow(data), floor(nrow(data)*0.50)) %>% 
    sort()
trainset

model_fit <- lm(Tmon_20110701 ~ x + y + Elevation_10km, data = data[trainset, ])

summary(model_fit) # very good!


# Evaluate with the trainig data
prediction <- predict(model_fit, newdata = data[-trainset, ])

cor(prediction, data$Tmon_20110701[-trainset])^2

```

With a model fitted, we can apply this to a raster stack. But before that, we need to get x and y coordinate rasters in the stack. We can *rasterize* a point data set for that.

```{r}

# xy_raster <- data %>% 
#     dplyr::mutate(x_coord = x,
#            y_coord = y) %>% 
#     terra::vect(geom = c("x_coord", "y_coord")) %>%
#     terra::rasterize(july2011, field = c("x", "y"))

temp <- stack
values(temp) <- 1
coords <- terra::crds(temp)
head(coords)

x_raster <- matrix(coords[,1], ncol = ncol(temp), byrow=TRUE) %>% 
    terra::rast() 
plot(x_raster)

y_raster <- matrix(coords[,2], ncol = ncol(temp), byrow=TRUE) %>% 
    terra::rast()
plot(y_raster)
x_raster

crs(x_raster) <- crs(stack)
crs(y_raster) <- crs(stack)

ext(x_raster) <- ext(stack)
ext(y_raster) <- ext(stack)
x_raster

xy_raster <- c(x_raster, y_raster)
names(xy_raster) <- c("x", "y")



prediction_stack <- c(xy_raster, stack)
prediction_stack
prediction_raster <- terra::predict(prediction_stack, model_fit)


plot(c(prediction_raster, july2011))

```




### Interpolating model to a raster


In the above examples we used data directly from rasters to fit a model and run it. We might not always have that luxury. Instead, we might have observations from certain locations, and need to fit our model using those points.

To pretend this kind of situation, we first randomise some points around Finland, and interpolate the model outputs using those points.

```{r}

# create random points
random_points <- terra::spatSample(july2011, 
                                   100, 
                                   method = "random", 
                                   as.points = TRUE)

plot(elevation)
points(random_points)

# extract data and fit a model
fitting_data <- terra::extract(stack, random_points, xy = TRUE) %>% 
    dplyr::filter(!is.nan(Tmon_20110701))
fitting_data

model_fit_points <- lm(Tmon_20110701 ~ x + y + Elevation_10km, 
                       data = fitting_data)
summary(model_fit_points)
stack

prediction_interpolated <- terra::interpolate(stack, model_fit_points)

plot(prediction_interpolated)
points(random_points)


plot(c(prediction_interpolated, july2011))
```



## Exercises


1.

* add a new raster layer *lakes* to prediction stack
* train a linear regression model with the extra explanatory variable
* **predict**
* plot

```{r}


```


2. 

* Extract values from prediction raster, and july2011
* compute correlation and mean error (*pred - obs*)
* plot a scatterplot with july2011 values on x-axis, and error on y-axis.

```{r}

```

