---
title: "CSC_D3S1 - Raster data in R"
author: "Marko Kallio"
date: "3 Nov 2021"
output: html_document
---



Let's make sure we have everything we need for Day 3.

```{r}
inst_packages <- installed.packages()
for (package in c("raster", "terra", "paletteer", "sf", "dplyr", "elsa",
                  "fasterize", "tmap", "classInt", "tidyr", "ggplot2")) {
    test <- package %in% inst_packages[,1]
    if (!test) {
        install.packages(package)
    }
}
rm(package, inst_packages, test)
```



## Raster basics using Terra

Let's first create two rasters from scratch.

```{r create matrix template}
library(terra)
library(paletteer)

side <- 15
empty_mat <- matrix(0, nrow = side, ncol = side)
empty_mat

mat1 <- empty_mat
for(i in 1:side) {
    mat1[,i] <- i
}
# look at the matrix, convert it to a raster and plot the raster
mat1
r1 <- terra::rast(t(mat1))
r1
terra::ext(r1) <- c(0,side,0,side) # set extent
r1
plot(r1)
```


Let's do another kind of raster, this time filled with random numbers.

```{r}
mat2 <- matrix(runif(length(empty_mat),0,10), nrow=side, ncol=side)
r2 <- terra::rast(mat2)
terra::ext(r2) <- c(0,side,0,side) # set extent
plot(r2, col = paletteer::paletteer_c("grDevices::Viridis", 100))
```


We now have two kind of rasters, one with gradually increasing values, and another one with random values. Let's compute Moran's I for both. {elsa} package works with objects from {raster} rather than {terra}, so we first convert.

```{r}
library(elsa)
r1 <- raster::raster(r1)
r2 <- raster::raster(r2)

elsa::moran(r1, d1 = 0, d2 = 10)
elsa::moran(r2, dl = 0, d2 = 10)
```
Just as expected. {elsa} has functions for a number of other spatial autocorrelation options too.

The raster values can be easily modified and stacked.

```{r}

r1 <- terra::rast(r1) # convert back to terra

r1 + 5
sqrt(r1)
r1^2
1 / r1

stack <- c(r1, r1+5, r1^2, 1/r1)
stack
names(stack) <- c("add 5", "square root", "square", "division")
plot(stack)

```




## Real world data

Let's read in the elevation model of Finland included in the data and inspect it.

```{r}
# for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"
elevation <- terra::rast("../Data/raster/Elevation_10km.tif")

summary(elevation)
ext(elevation) # this is the bounding box of the raster
plot(elevation, col = paletteer::paletteer_c("viridis::viridis", 100))

elevation[elevation == 0] <- NA
plot(elevation, col = paletteer::paletteer_c("viridis::viridis", 100))
summary(elevation)
```

We can compute statistics from the elevation layer using *global()* function.

```{r}

global(elevation, mean, na.rm=TRUE)
global(elevation, median, na.rm=TRUE)
global(elevation, range, na.rm=TRUE)
global(elevation, sd, na.rm=TRUE) 

plot(stack)
global(stack, mean)

```


## Exercises

1.
remember: for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"

* Read in a temperature layer from "../Data/raster/Tmean_Monthly_2005-2014/Tmon_20050101.tif"
* Read in two other layers from the same folder, but from a different month. The number in the file refers to YYYYMMDD.
* Stack the three layers in one
* compute mean temperature
* plot a histogram using *hist()*

```{r}
t01 <- rast("../Data/raster/Tmean_Monthly_2005-2014/Tmon_20050101.tif")
t07 <- rast("../Data/raster/Tmean_Monthly_2005-2014/Tmon_20050701.tif")
t10 <- rast("../Data/raster/Tmean_Monthly_2005-2014/Tmon_20051001.tif")

t <- c(t01, t07, t10)

global(t, mean, na.rm=TRUE)



```


2.

* plot the temperature layers
* use a colour palette using paletteer_c()


```{r}
plot(t)

brk <- seq(-11,20, length.out = 10)
brk

plot(t, col = paletteer_c("scico::turku", 100), breaks = brk)


```

