---
title: "CSC_D3S2 - Raster data manipulation and extraction"
author: "Marko Kallio"
date: "3 Nov 2021"
output: html_document
---

## Raster data manipulation




```{r}
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(tmap)

# for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"
admin <- sf::read_sf("../Data/raster/SuomenKuntajako_2021_100k.shp")
plot(admin)

helsinki <- dplyr::filter(admin, NAMEFIN == "Helsinki")
plot(admin)

plot(helsinki)

```


Read in a list of files.

```{r}
# for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"
files <- list.files("../Data/raster/Tmean_Monthly_2005-2014/", 
                    full.names = TRUE)
files

temperature <- lapply(files, terra::rast) #read in all files, output is a list

length(temperature)
temperature[[1]]
temperature[1:5]


temperature <- do.call("c", temperature) # stack all rasters together
temperature

plot(temperature)


temperature <- terra::rast(files)
plot(temperature)

```


Before we do anything, let's make sure both datasets are in the same coordinate reference system. We'll use the CRS of temperature raster stack, because tranforming a raster changes its values.

```{r}
( sfcrs <- sf::st_crs(helsinki) )
terra::crs(temperature)
class(sfcrs)
str(sfcrs)
sfcrs$wkt

helsinki <- sf::st_transform(helsinki, terra::crs(temperature))

st_crs(helsinki)


# we could project the raster with:
temperature <- terra::project(temperature, sf::st_crs(helsinki)$wkt)

temp <- temperature[[1]]
temp
plot(temp)

temp_projected <- project(temp, "epsg:4326" )
temp_reprojected <- project(temp_projected, temp )

plot( c(temp, temp_reprojected) )
c(temp, temp_reprojected)

all.equal(temp, temp_reprojected)

plot(temp_projected)
plot(temp_reprojected)
```


### Extracting values using a polygon

Now we can find out all the values in the raster stack for Helsinki! to do this in {terra}, we need to convert the sf-helsinki into a SpatVector-helsinki. This works the same way for lines and points as well.

```{r}
helsinki_terra <- terra::vect(helsinki)
helsinki_temps <- terra::extract(temperature, helsinki_terra)
helsinki_temps


helsinki_temps <- dplyr::summarise_all(helsinki_temps, mean, na.rm=TRUE) %>% 
    tidyr::gather(date, temperature, -ID) %>% 
    dplyr::mutate(date = stringr::word(date, 2, sep = "_"))

helsinki_temps
plot(helsinki_temps[,3])
lines(helsinki_temps[,3])

```


### Aggregation and resampling

Aggregation is done by providing the function a factor - how many cells in each direction are used to aggregate. The temperature has a resolution of 10km. Aggregating by factor 4 yields a raster with resolution of 40km (by 40km).

```{r}
temp_40km <- terra::aggregate(temperature, fact = 4, fun = mean, na.rm=TRUE)
plot(temp_40km)
temp_40km

```

We can also disaggregate - but the original values cannot be returned!

```{r}
# temp_10km <- terra::disagg(temp_40km, fact = 4, method = "bilinear")
temp_10km <- disaggregate(temp_40km, fact = 4, method = "bilinear")
temp_10km

temp_difference <- temp_10km - temperature

plot(temp_difference[[1]])
```

Resampling can achieve both. But are they the same - no!

```{r}

temp_40km_resampled <- terra::resample(temperature, temp_40km, method = "bilinear")
plot(temp_40km_resampled)
temp_40km_resampled
temp_40km
plot(temp_40km - temp_40km_resampled)

```


### Spatial subsets via cropping

Let's have a closer look at temperatures in Turku. First we'll get a bounding box of the municipality, and crop the raster using that.

```{r}

Turku_temps <- terra::crop(temperature,
                           admin %>% 
                               dplyr::filter(NAMEFIN == "Turku") %>% 
                               terra::vect(),
                           snap = "out")

plot(Turku_temps[[1]])
plot(sf::st_geometry(admin), add=TRUE)


```


## focal statistics,

or the so-called moving window statistics. We can use any function for focal statistics, as long as it takes an input of multiple numerical values, and returns a single value only.
However, it only runs for a single layer
```{r}

# ntice the message saying it only used the first layer
temperature_smoothed <- terra::focal(temperature,
                                     w = 21,
                                     fun = max)

plot(temperature_smoothed)
plot(temperature[[1]])

```

We can also define the neighbourhood in other ways:
```{r}
temperature_smoothed <- terra::focal(temperature[[1]],
                                     w = c(nrow(temperature)-1, 3),
                                     fun = mean)
plot(temperature_smoothed)

# remove values from the sea
temperature_smoothed[is.na(temperature[[1]])] <- NA
plot(temperature_smoothed)


```

or with matrices:

```{r}

weight_matrix <- matrix(1, nrow = nrow(temperature)-1, ncol = 1)
weight_matrix

temperature_smoothed <- terra::focal(temperature[[1]],
                                     w = weight_matrix,
                                     fun = mean,
                                     na.rm=TRUE)

plot(temperature_smoothed)

# remove values from the sea
temperature_smoothed[is.na(temperature[[1]])] <- NA
plot(temperature_smoothed)

```


### reclassification

The final part of this lesson shows reclassification. Here we divide Finland into classes based on mean temperature.

But before that, lets compute the mean temeprature at each cell over the 10-year period 2005-2014, and find "optimal" class intervals.

```{r}
plot(temperature)

temperature_mean <- mean(temperature)
temperature_mean
plot(temperature_mean)



library(classInt)

intervals <- classInt::classIntervals(terra::values(temperature_mean),
                            n = 4,
                            style = "quantile")

# quantile(values(temperature_mean), probs = c(0.2, 0.4, 0.6, 0.8), na.rm=TRUE)

intervals  
         
reclass_matrix <- matrix(c(-Inf, intervals$brks[1], 1,
                           intervals$brks[1], intervals$brks[2], 2,
                           intervals$brks[2], intervals$brks[3], 3,
                           intervals$brks[3], intervals$brks[4], 4,
                           intervals$brks[4], Inf, 5),
                         ncol = 3,
                         byrow = TRUE)
reclass_matrix

temperature_classified <- terra::classify(temperature_mean,
                                          reclass_matrix)

plot(temperature_classified)

```


And finally, let's save that raster for later use.
```{r}
# for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"
terra::writeRaster(temperature_classified, 
                   filename = "../Data/raster/temperature_classified.tif",
                   overwrite = TRUE)


```




## Exercises

1.

* Choose a municipality from the municipality dataset (larger municipalities are better). Crop the temperature raster stack to that municipality.
* find the mean, min, and max temperatures for that municipality over the period 2005-2014 (the full raster stack)

```{r}
kolari <- admin %>% 
    filter(NAMEFIN == "Kolari")

temp_kolari <- crop(temperature, kolari)

values <- global(temp_kolari, mean) 
mean_temp <- mean(values$mean)


values <- terra::extract(temp_kolari, vect(kolari))
mean(as.matrix(values[,-1]),na.rm=TRUE)


```


2. 

* compute a focal function over the temperature_mean raster with the window of your choosing (square, rectangular, or some other matrix. Look at the help for *terra::focal()*. Try e.g. the Laplacian or Sobel filters?)
* Crop both rasters to the municipality you chose: the temperature_mean raster and the focal raster you've just created.
* Plot the rasters - do you see large differences? Compute the difference between the rasters, and plot that too.


```{r}
focal_mean <- focal(temperature_mean, 
                    w = matrix(c(0,1,0,1,-4,1,0,1,0), nrow=3))

plot(focal_mean)

focal_mean <- crop(focal_mean, kolari)
plot(c(focal_mean, mean(temp_kolari)))

plot(focal_mean - mean(temp_kolari))

```


3. 

* Reclassify both rasters using just two classes: *hot* and *cold*. Where are the hot and cold parts of the municipality?


```{r}

reclass_matrix <- matrix( c(-Inf, mean(values(temp_kolari), na.rm=TRUE), factor("cold", levels = c("hot", "cold")),
                            mean(values(temp_kolari), na.rm=TRUE), Inf, factor("hot", levels = c("hot", "cold"))),
                          ncol = 3,
                          byrow = TRUE)

reclass_matrix

temp_kolari_classified <- classify(mean(temp_kolari),
                        reclass_matrix)

plot(mean(temp_kolari))
plot(temp_kolari_classified)
```


