---
title: "CSC_D3S3 - Map algebra"
author: "Marko Kallio"
date: "12 May 2023"
output: html_document
---


We have already looked at most of Map algebra already - local, focal and global functions. Here we will look at these a little bit more in detail, and include zonal functions here too.

```{r}
library(terra)
library(dplyr)

files <- list.files("../Data/raster/RRmon_Monthly_2005-2014/", 
                    full.names = TRUE, pattern = "*.tif$")

rainfall <- lapply(files, terra::rast) #read in all files, output is a list
files
```

Oops! There are some non-raster files in the file list with the extension ".tif.aux.xml"! These are created e.g. when the raster file is opened in a Desktop GIS. Let's get rid of them and try again!


```{r}
files <- list.files("../Data/raster/RRmon_Monthly_2005-2014/", 
                    full.names = TRUE)
files <- files[!grepl(".xml", files)] # deselect all files which contain string ".xml"
files

# rainfall <- lapply(files, terra::rast) #read in all files, output is a list
# rainfall <- do.call("c", rainfall) # stack all rasters together
rainfall <- terra::rast(files)
plot(rainfall[[1]])
```



## Local functions

Local functions are those which deal with single cells at the time, without considering any neighbours. Local functions can be applied directly to the {terra} SpatRasters.

```{r}

rainfall / 30 # approximate per-day rainfall, assuming 30-day months

mean(rainfall) # average per-cell rainfall

max(rainfall) # maximum per-cell rainfall

rainfall_quantiles <- quantile(rainfall, 
                               probs = c(0.05, 0.5, 0.95)) # 5th, 50th, and 95th percentiles

plot(rainfall_quantiles)
```

We can also use our own defined functions through *terra::app()*. Here we compute the coefficient of variation for rainfall.

```{r}
cv <- function(x, na.rm=FALSE) {
    stdev <- sd(x, na.rm=na.rm)
    mn <- mean(x, na.rm=na.rm)
    
    return(stdev/mn)
}

# cv(rainfall) #this fails!

rainfall_cv <- terra::app(rainfall, fun = cv)

plot(rainfall_cv)

```


## focal functions

We have looked at focal functions quickly over the last session, and made some exercises. *terra::focal()* can take user defined functions as well. Here we compute coefficient of variation over a window of 30km for the first layer of rainfall. 

NOTE 2023-05-06: The behavior of focal() has changed, and if input a SpatRaster with multiple layers, output is corrupted. 

```{r}

rainfall_cv_50km <- terra::focal(rainfall[[1]],
                                 w = 5,
                                 fun = cv,
                                 na.rm=TRUE)

plot(rainfall_cv_50km)

```

Focal does not (yet) support multiple layers, but we can overcome this issue with some tricks:

```{r}

rainfall_cv <- terra::focal(rainfall,
                            w = 5,
                            fun = cv,
                            na.rm=TRUE)
plot(rainfall_cv)



rainfall_cv_2 <- lapply(1:12, function(i) {
    cv_50km <- terra::focal(rainfall[[i]], w = 5, fun = cv)
    return(cv_50km)
}) %>% do.call("c", .)

plot(rainfall_cv_2)
```



### Zonal functions

Zonal functions have not been addressed yet. In zonal operations, some function is applied to all the values within specified zones. For an example, we will use the classified temperature zones we created at the end of last session. We'll find the mean rainfall falling within those zones.

```{r}

temperature_zones <- rast("../Data/raster/temperature_classified.tif")
plot(temperature_zones)

( zonal_rainfall <- terra::subset(rainfall, 1:60) %>% 
        terra::zonal(temperature_zones, fun = mean) )

library(ggplot2)
library(tidyr)
library(lubridate)
library(stringr)

data <- tidyr::gather(zonal_rainfall, date, rainfall, -mean) %>% 
    mutate(date = lubridate::as_date(stringr::word(date, 2, sep = "_" )),
           zone = factor(mean))
data

ggplot(data) +
    geom_line(aes(date, rainfall, color = zone)) +
    paletteer::scale_color_paletteer_d(palette = "lisa::Rembrandt")

```

What if we'd like to know the mean rainfall in each of Finland's municipalities? first we'd need to rasterize the admin boundaries, and apply zonal operation to them.

```{r}

admin <- terra::vect("../Data/raster/SuomenKuntajako_2021_100k.shp")
admin
admin$NATCODE <- as.numeric(admin$NATCODE)

admin_raster <- terra::rasterize(admin,
                                 rainfall, # this is the target raster which is used for rasterization
                                 field = "NATCODE") # which attribute we're rasterizing 

plot(admin_raster)

rainfall_admin <- terra::zonal(mean(rainfall), admin_raster, fun = mean, na.rm=TRUE)
rainfall_admin

rainfall_admin <- admin %>% 
    sf::st_as_sf() %>% 
    dplyr::select(NATCODE, NAMEFIN) %>% 
    dplyr::left_join(rainfall_admin, by = "NATCODE") %>%
    dplyr::arrange(-mean)

rainfall_admin

plot(rainfall_admin[,"mean"])
```



## Global functions

Global functions are those which are applied to all the values in a raster layer at once. These can be simple ones like computing the mean of the raster values, or more complex, like a viewshed function.

```{r}

terra::global(rainfall, fun = mean, na.rm=TRUE)
terra::global(rainfall, fun = quantile, probs = c(0.05, 0.5, 0.95), na.rm=TRUE)
terra::global(rainfall, fun = cv, na.rm=TRUE)



```



## Exercises

1.

* Read in one temperature layer from "../Data/raster/Tmean_monthly_2005-2014/".
* Convert the temperatures in that layer from Celcius to Kelvin.

```{r}
files <- list.files("../Data/raster/Tmean_Monthly_2005-2014/",
                    full.names = TRUE)
files
temp <- rast(files[2])

plot(temp + 273.15)

```

2. 

* Create a zone (rasterize)for a munucipality of your choosing (bigger ones better - use only that municipality polygon when rasterizing)
* Compute some zonal statistics for that municipality.

```{r}
muonio <- admin[1,]
muonio

muonio_r <- rasterize(muonio, temp, field = "NATCODE")

plot(muonio_r)

zonal(temp, muonio_r, fun = "mean", na.rm=TRUE)

zonal(temp, vect(muonio), fun = "mean", na.rm=TRUE)

plot(as.polygons(muonio_r)) # raster -> polygon conversion
```

