---
title: 'CSC Spatial Analysis with R: Day 1, exercise answers'
author: "Marko Kallio"
date: "10 May 2023"
output:
  html_document:
    df_print: paged
    theme: flatly
---

# Exercises
## D3S1

1.

* Read in a temperature layer from "../Data/raster/Tmean_Monthly_2005-2014/Tmon_20050101.tif"
* Read in two other layers from the same folder, but from a different month. The number in the file refers to YYYYMMDD.
* Stack the three layers in one
* compute mean temperature
* plot a histogram using *hist()*

```{r}
library(terra)
library(dplyr)

temp1 <- rast("../Data/raster/Tmean_Monthly_2005-2014/Tmon_20050101.tif")
temp2 <- rast("../Data/raster/Tmean_Monthly_2005-2014/Tmon_20050201.tif")
temp3 <- rast("../Data/raster/Tmean_Monthly_2005-2014/Tmon_20050301.tif")

stack <- c(temp1, temp2, temp3)

plot(mean(stack))
global(stack, mean, na.rm=TRUE)

hist(stack)

?`hist,SpatRaster-method`
```


2.

* plot the temperature layers
* use a colour palette using paletteer_c()


```{r}
library(paletteer)

plot(stack)
plot(stack, col = paletteer_c(`"ggthemes::Red-Blue-White Diverging"`, 10, direction = -1))


summary(stack)
breaks <- seq(-12.5, 12.5, by = 5)
breaks
# colors <- paletteer::paletteer_c("scico::turku", 10)
colors <- paletteer_c(`"ggthemes::Red-Blue-White Diverging"`, 5, direction = -1)
plot(colors)

plot(stack, breaks = breaks, col = colors)


?`plot,SpatRaster,numeric-method`
```


## D3S2

1.

* Choose a municipality from the municipality dataset (larger municipalities are better). Crop the temperature raster stack to that municipality.
* find the mean, min, and max temperatures for that municipality over the period 2005-2014 (the full raster stack)

```{r}
admin <- sf::read_sf("../Data/raster/SuomenKuntajako_2021_100k.shp")

files <- list.files("../Data/raster/Tmean_Monthly_2005-2014/", 
                    full.names = TRUE)

temperature <- lapply(files, terra::rast) #read in all files, output is a list
temperature <- do.call("c", temperature) # stack all rasters together

inari <- admin %>% filter(NAMEFIN == "Inari")
plot(inari)
temp_inari <- crop(temperature, inari)

# Calculate the same thing in a few different ways.
new_func <- function(x, na.rm=FALSE) {
  mn <- mean(x, na.rm = na.rm)
  min <- min(x, na.rm = na.rm)
  max <- max(x, na.rm = na.rm)
  return(c(mean = mn, min = min, max = max))
}
global(temp_inari, fun = new_func, na.rm=TRUE)

# Requires newer version of Terra to work.
# global(temp_inari, c("mean", "min", "max"), na.rm=TRUE)

global(temp_inari, mean, na.rm=TRUE)
global(temp_inari, min, na.rm=TRUE)
global(temp_inari, max, na.rm=TRUE)

mean_inari <- mean(temp_inari)
plot(mean_inari)
```

2. 

* compute a focal function over the temperature_mean raster with the window of your choosing (square, rectangular, or some other matrix. Look at the help for *terra::focal()*. Try e.g. the Laplacian or Sobel filters?)
* Crop both rasters to the municipality you chose: the temperature_mean raster and the focal raster you've just created.
* Plot the rasters - do you see large differences? Compute the difference between the rasters, and plot that too.

```{r}
focal_mean <- focal(mean_inari, w = 3, fun = mean)

inari_stack <- c(mean_inari, focal_mean)
plot(inari_stack)
```


3. 

* Reclassify both rasters using just two classes: *hot* and *cold*. Where are the hot and cold parts of the municipality?

```{r}
mean_temp <- global(mean_inari, mean, na.rm=TRUE)
mean_temp

reclass_mat <- matrix(c(-Inf, mean_temp$mean, 0,
                        mean_temp$mean, Inf, 1),
                         ncol = 3,
                         byrow = TRUE)
reclass_mat
hotcold_inari <- terra::classify(mean_inari, rcl = reclass_mat)
plot(hotcold_inari)

```

## D3S3

1.

* Read in one temperature layer from "../Data/raster/Tmean_monthly_2005-2014/".
* Convert the temperatures in that layer from Celcius to Kelvin.

```{r}
files <- list.files("../Data/raster/Tmean_Monthly_2005-2014/",
                    full.names = TRUE)
files
temp <- rast(files[2])

plot(temp + 273.15)

```

2. 

* Create a zone (rasterize)for a munucipality of your choosing (bigger ones better - use only that municipality polygon when rasterizing)
* Compute some zonal statistics for that municipality.

```{r}
muonio <- admin[1,]
muonio

muonio_r <- rasterize(muonio, temp, field = "NATCODE")

plot(muonio_r)

zonal(temp, muonio_r, fun = "mean", na.rm=TRUE)

# Alternative option, to use vector data with zonal function
zonal(temp, vect(muonio), fun = "mean", na.rm=TRUE)

plot(as.polygons(muonio_r)) # raster -> polygon conversion
```



## D3S4

1.

* add a new raster layer *lakes* to prediction stack
* train a linear regression model with the extra explanatory variable
* **predict**
* plot

```{r}
# Variables needed from earlier code:

elevation <- terra::rast("../Data/raster/Elevation_10km.tif") # Digital elevation model
lakes <- terra::rast("../Data/raster/Lake_10km.tif") # Lake cover

july2011 <- temperature[["Tmon_20110701"]]
july2011 <- terra::project(july2011, elevation, method = "near")

stack <- c(july2011, elevation)

data <- terra::as.data.frame(stack, xy = TRUE)

xy_raster <- data %>% 
  dplyr::mutate(x_coord = x, y_coord = y) %>% 
  terra::vect(geom = c("x_coord", "y_coord")) %>%
  terra::rasterize(july2011, field = c("x", "y"))

prediction_stack <- c(xy_raster, stack)

trainset <- sample(1:nrow(data), floor(nrow(data)*0.50)) %>% 
    sort()

model_fit <- lm(Tmon_20110701 ~ x + y + Elevation_10km, data = data[trainset, ])
prediction_raster <- terra::predict(prediction_stack, model_fit)
```

```{r}
# Exercise part:

prediction_stack <- c(prediction_stack, lakes)
plot(prediction_stack)

lakes_data <- terra::as.data.frame(prediction_stack, xy = TRUE)

lakes_model_fit <- lm(Tmon_20110701 ~ x + y + Elevation_10km + Lake_10km, data = lakes_data[trainset, ])

summary(model_fit)
summary(lakes_model_fit)
lakes_prediction_raster <- terra::predict(prediction_stack, lakes_model_fit)

plot(c(no_lakes = prediction_raster, lakes = lakes_prediction_raster, july2011))


```


2. 

* Extract values from prediction raster, and july2011
* compute correlation and mean error (*pred - obs*)
* plot a scatterplot with july2011 values on x-axis, and error on y-axis.

```{r}

predict_data <- as.data.frame(lakes_prediction_raster, xy = TRUE)
observed_data <- as.data.frame(july2011, xy = TRUE)

task2_data <- dplyr::left_join(predict_data, observed_data, by = c("x", "y")) 
task2_data

cor(task2_data$lyr1, task2_data$Tmon_20110701, use = "complete.obs")
error <- task2_data$lyr1 - task2_data$Tmon_20110701

mean(error, na.rm=TRUE)

plot(task2_data$Tmon_20110701, error)
```

