---
title: "CSC Spatial Analysis with R: Day 1 Session 2 - Plotting spatial data"
author: "Marko Kallio"
date: "1 Nov 2021"
output: html_document
---


Spatial analysis depends on our ability to visualize data, create maps and use them to *see* patterns which often cannot picked up by numerical analysis. In this session we look at basic plotting of {sf} objects using both base R plot() function, as well as ggplot2.



## Base R plot()

```{r}
library(sf)

# recreate the objects from last session
point <- st_point(c(0,0))

coord_matrix <- rbind(c(1.5,1.5), c(1.5,2.5), c(2.5,2.5), c(2.5,1.5))
line <- st_linestring(coord_matrix)

polygon_border <- rbind(c(0,0), c(0,3), c(3,3), c(3,0), c(0,0))
polygon_hole <- rbind(c(1,1), c(1,2), c(2,2), c(2,1), c(1,1))
polygon_with_hole <- st_polygon( list(polygon_border, polygon_hole))

plot(point)
plot(line)
plot(polygon_with_hole)
```

We can include them all in the same plot.
```{r}
plot(polygon_with_hole)
plot(line, add=TRUE)
plot(point, add=TRUE)
```

And we can control their aestethics in the same way as any base R plot.
```{r}
plot(polygon_with_hole, col='blue')
plot(line, col='red', lwd=5, add=TRUE)
plot(line, col='black', lty=2, lwd=3, add=TRUE)
plot(point, col='black',  pch=16, cex=3, add=TRUE)
```

Now let's try with some real data. 
```{r}
# for CSC notebooks
# villages <- sf::read_sf("/home/rstudio/r-spatial-course/Data/villages.gpkg")

# for windows
villages <- sf::read_sf("../data/villages.gpkg")

plot(villages)
plot(villages, max.plot=3)
plot(villages[,"PovertyRate"])
plot(villages[,"prov_name"])
plot(sf::st_geometry(villages))
```

There are a large number of options available to control the plot. To find them all, type help("plot.sf") to the console. Here's an example of adding a graticule.
```{r}
plot(villages[,"LiteracyRate"], graticule = TRUE)
```

As with the geometry types, we can add many layers of {sf} objects to the plot. Here we read provincial borders of Laos and add them to the plot. 

Notice that when plotting {sf} objects, we must add parameter *reset = FALSE* in order to add new layers.
```{r}
# for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"
# admin <- sf::read_sf("/home/rstudio/r-spatial-course/data/LAO_adm1.shp")
admin <- sf::read_sf("../data/LAO_adm1.shp")

plot(villages[,10], reset=FALSE)
plot(sf::st_geometry(admin), add=TRUE)
```



## Plotting with ggplot2

One of the strengths of R is the wonderful visualization package *ggplot2*. We can also plot {sf} objects with it using *geom_sf()*. There are small differences when compared to other geom functions. First, aestethics are not required, because those come directly from the geometry. Second, the data must be directly given to *geom_sf*.

```{r}
library(ggplot2)
ggplot() + geom_sf(data = villages, size = 0.2)

```

However, aestethics are useful for colours.
```{r}
ggplot() + 
  geom_sf(data = admin) +
  geom_sf(data = villages, aes(color = PovertyRate))

ggplot() + 
  geom_sf(data = admin, aes(fill=NAME_1)) +
  geom_sf(data = villages, size = 0.2)
```

*ggplot2* is extremely powerful, and we can control pretty much anything in the plot.
```{r}
ggplot() + 
  geom_sf(data = admin, fill='grey50') + 
  geom_sf(data = villages, aes(color = ToiletType), show.legend = FALSE) + # no legend
  theme_bw() + # black and white theme for background
  labs(title="Villages in Laos", x = "Longitude", y = "Latitude") + # title and axis labels
  scale_colour_viridis_c()
```

Removing graticule is possible only if removed **after** *geom_sf()*. We can zoom in, too.
```{r}
ggplot() + 
  geom_sf(data = admin, fill='grey50') + 
  geom_sf(data = villages, aes(color = ToiletType), show.legend = FALSE) + # no legend
  theme_bw() + # black and white theme for background
  labs(title="Villages in Laos", x = "Longitude", y = "Latitude") + # title and axis labels
  scale_colour_viridis_c() +
  coord_sf(datum=NA, xlim = c(103,105), ylim = c(20,22)) # remove graticule and control plot extent
```

The plot, however, also lost the latitude and longitude ticks. We can, also, just make the graticule transparent.

```{r}
ggplot() + 
  geom_sf(data = admin, fill='grey50') + 
  geom_sf(data = villages, aes(color = ToiletType)) + 
  theme_bw() + # black and white theme for background
  labs(title="Villages in Laos", x = "Longitude", y = "Latitude") + # title and axis labels
  scale_colour_viridis_c() +
  theme(panel.grid.major=element_line(colour="transparent")) + # make graticule transparent
  coord_sf(xlim = c(103,105), ylim = c(20,22)) # control plot extent
```

Several plots can also be combined, with {patchwork} like here where we plot a map and accompanying scatterplot and a boxplot. 
```{r}
library("patchwork")

p1 <- ggplot() + 
  geom_sf(data = admin, fill='grey50') + 
  geom_sf(data = villages, aes(color = IrrigationType)) + 
  theme_bw() + # black and white theme for background
  labs(title="Villages in Laos", x = "Longitude", y = "Latitude") + # title and axis labels
  scale_colour_viridis_c() +
  theme(panel.grid.major=element_line(colour="transparent")) + # make graticule transparent
  coord_sf(xlim = c(103,105), ylim = c(20,22)) # control plot extent

p2 <- ggplot(villages, aes(x = PovertyRate, y = LiteracyRate)) +
  geom_point() +
  geom_smooth()

p3 <- ggplot(villages, aes(x = prov_name, y=PovertyRate)) +
  geom_boxplot()

p1 | (p2/p3)

```


## Interactive mapping using 'mapview'

One of the main issues in using R for GIS is the lack of interactivity. However, recent new packages have considerably improved R's capabilities in this sector. One such package is 'mapview', which automatically creates a Leaflet-map from a large list of input classes (including {sf}, 'sp', and 'raster'). Usage of mapview is simple:

```{r include=FALSE}
library(mapview)

mapview(villages)
```

We can add additional layers to the same view..

```{r}
mapview(admin) + villages
```

... and control their aestethics.

```{r}
mapview(admin, zcol = "NAME_1") + mapview(villages, zcol = "IrrigationType")
```


Interactive selection and editing can be done with package 'mapedit', but we will not go through the package in this course. See the package github page for more information.


## Exercises

1. 

* Create a ggplot which shows admin areas, villages and rivers.
* visualize the rivers as dashed line using parameter *lty = 2* in the geom_sf() function call.
* give the admin areas a white or transparent *fill* colour.
* reduce the *size* of the village dots

```{r}
# for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"
villages <- read_sf("../data/villages.gpkg")
admin <- read_sf("../data/LAO_adm1.shp")
rivers <- read_sf("../data/rivers.gpkg")

```

2. Create another map zooming into one full province. Hint: Use *filter* to pick one province, and get the bounding box using function *st_bbox()*.

```{r}

```

