---
title: "CSC Spatial Analysis with R: Day 1 Session 3 - Vector data manipulation"
author: "Marko Kallio"
date: "1 Nov 2021"
output:
  html_document:
    df_print: paged
    theme: flatly
---


This lesson of the day spatial and geometric operations. In this lesson you'll learn how to:

* transform between geometry types
* handle coordinate reference systems in R
* do operations on object geometries
* do spatial operations (which do not affect object geometries)


# Geometric operations

The file basins.gpkg does not include information on the coordinate reference system, and hence we need to set it (we know it is in EPSG:4326). Basins is a small subset of the HydroBasins dataset (http://hydrosheds.org) including all those basins which intersect Laos.

```{r message=FALSE, warning=FALSE, results='hide'}
library(sf)
library(dplyr)
library(mapview)
```

```{r message=FALSE, warning=FALSE}
# for CSC notebooks "/home/rstudio/r-spatial-course/"
# basins <- sf::read_sf("/home/rstudio/r-spatial-course/Data/basins.gpkg") %>%
#     sf::st_set_crs(4326)
basins <- sf::read_sf("../../Data/basins.gpkg") %>%
    sf::st_set_crs(4326)

plot(basins, max.plot=3)
mapview::mapview(basins)
```


## Transforming object types

To transform polygons to different object representations (a line, or a point), we can use the function st_cast(). But before that, we union polygons by attribute (equivalent to Dissolve in QGIS) polygons using their attribute MAIN_BAS. In sf, dissolving happens by grouping and summarising. Note that we can do operations while summarising too: here we find the total area of the dissolved polygons.

```{r message=FALSE, warning=FALSE}
major_basins <- basins %>% 
    dplyr::group_by(MAIN_BAS) %>% 
    dplyr::summarize(total_area = sum(SUB_AREA))

major_basins_outline <- sf::st_cast(major_basins, "LINESTRING")
major_basins_nodes <- sf::st_cast(major_basins, "MULTIPOINT")

plot(major_basins)
plot(major_basins_outline)
plot(major_basins_nodes)
```




The MULTIPOINT features can also be coerced into linestrings or polygons. However, for complex features this may be difficult. Below, some islands are joined to the mainland.
```{r message=FALSE, warning=FALSE}
major_basins_outline <- sf::st_cast(major_basins_nodes, "LINESTRING")
plot(major_basins_outline)

major_basins_nodes %>%
  sf::st_cast("POLYGON") %>% 
  plot()


# POINTs cannot be coerced directly to LINESTRINGS or POLYGONS. The following produces an error
#major_basins_nodes <- st_cast(major_basins, "POINT")
#major_basins <- st_cast(major_basins_nodes, "POLYGON")
```




There are still some packages around which do not support {sf}, but use the older {sp} package. Conversions between these two may need to be done:

```{r}
basins_sp <- as(basins, "Spatial")
basins_sf <- sf::st_as_sf(basins_sp)
```


## Coordinate reference systems

Unlike many graphical GIS programs (e.g. ArcGIS, QGIS), R cannot transform objects on the fly. Therefore, having all data in the same CRS is important. Luckily, functions in {sf} package will give an error if the CRS of input objects is not identical. Inspecting the CRS of an object can be done with *st_crs()* function. Here we load "rivers.gpkg", which is a subset of HydroSheds river network, derived from 15 arc second DEM.

```{r}
# for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"
basins <- sf::read_sf("../Data/basins.gpkg", quiet = TRUE)
rivers <- sf::read_sf("../Data/rivers.gpkg", quiet = TRUE)
sf::st_crs(basins)
sf::st_crs(rivers)

```




If we know the CRS of our data, but it is missing from the object, we can set it using *st_set_crs()* function. Note that **this will not transform the coordinates!** Setting can be easily done either by using the EPSG code or proj.4 string. 

```{r message=FALSE, warning=FALSE}
# using EPSG SRID
basins <- sf::st_set_crs(basins, 4326)

# using proj.4 string
basins <- sf::st_set_crs(basins, "+init=epsg:4326")
# basins <- sf::st_set_crs(basins, "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")

# OR
sf::st_crs(basins) <- 4326
# sf::st_crs(basins) <- "+init=epsg:4326"
# sf::st_crs(basins) <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
```




Coordinate transformations are likewise easy to achieve using *st_transform()* function. The input can be either the EPSG or proj.4 string, as with *st_set_crs()*. Here we transform basins to the same CRS as rivers (UTM zone 48N; EPSG:32648).

```{r}
basins <- sf::st_transform(basins, 32648) 

# can also transform directly to the CRS of another object:
# basins <- sf::st_transform(basins, st_crs(rivers)) 

sf::st_crs(basins)
```


## Simplification, centroids, buffers, clipping, intersection...

Geometric operations include all operations which alter geometries, or result in new ones. Such commands include for instance simplifications, clipping, difference, intersection, computing centroids, buffers etc. {sf} uses the OGC specifications for these operations, and therefore differ slightly from the terminology used in e.g. ArcGIS or QGIS.


### Union (Dissolve)
Here we create major basins from smaller ones by creating a union of the small geometries based on their attribute "MAIN_BAS". This is equivalent to dissolving in ArcGIS and QGIS. There is no ready made function in {sf} for this, but we can use the predicates provided in 'dplyr' package. Function st_union() from {sf} creates a union of all objects in the input geometry (equals dissolving without attributes in ArcGIS or QGIS).

```{r message=FALSE, warning=FALSE}
# The provided st_union() function creates a union of all geometries in the input
( major_basins <- st_union(basins) )

# dissolve to major basins again:
major_basins <- basins %>% 
    dplyr::group_by(MAIN_BAS) %>% 
    dplyr::summarize()
plot(major_basins)

```





### Simplify

With the dissolved major basins, we can try simplification using st_simplify(). It operates on each individual geometry separately, and thus does not preserve uniform polygon borders. For polygon feature sets which share borders, we could use function ms_simplify() from 'rmapshaper' package.

```{r message=FALSE, warning=FALSE}
# st_simplify() takes toleranse in the units of CRS. In our case it is UTM, so the unit is meters.
simplified_st <- sf::st_simplify(major_basins, dTolerance = 5000)
plot(simplified_st)

# st_simplify() 
library(rmapshaper, quietly = TRUE)
# keep 1% of nodes. keep_shapes prevents small polygons from disappearing at high simplification
simplified_ms <- rmapshaper::ms_simplify(major_basins, keep = 0.01, keep_shapes=TRUE)
plot(simplified_ms)
```





### Centroids

```{r message=FALSE, warning=FALSE}
centroids_sf <- sf::st_centroid(basins)
plot(dplyr::select(basins, UP_AREA), reset = FALSE) # Here using the dplyr::, because also raster package has a function named select(), and it creates conflict. the dplyr:: specifies that it is looking for the function in package 'dplyr'.
plot(centroids_sf, col='white', pch=16, add=TRUE)
```





#### Buffers

Buffers can be taken from any {sf} object. Here we take a fixed buffer of 5km from the centroid of each feature in major_basins.
```{r message=FALSE, warning=FALSE}
buffer_sf <- major_basins %>%
    sf::st_centroid() %>% 
    sf::st_buffer(dist = 15000)

plot(buffer_sf)
```




Variable buffer are also possible by giving the input argument dist a vector of the same length as there are features in the {sf} object. Here we take a random buffer 2-10km wide.

```{r message=FALSE, warning=FALSE}
# random buffer distances
buf_dist <- nrow(major_basins) %>% 
    runif(10000, 30000)

# sf
varbuf_sf <- major_basins %>%
    sf::st_centroid() %>% 
    sf::st_buffer(dist = buf_dist)


plot(varbuf_sf, reset = FALSE)
major_basins %>%
    sf::st_cast("LINESTRING") %>%
    plot(col='black', add=TRUE)
```


### Creating regular grids (fishnets)

Fishnets or regular grids of points can be created with *st_make_grid()* function. It creates a regular grid over bounding box of an {sf} object. Can be given a certain cellsize, or number of cells in x and y directions. 'what' tells the function what kind of regular grid is wanted ('polygons', or points in 'corners', or 'centers'). Fishnets of lines rather than polygons can be created simply by casting the polygons as "LINESTRING"s.
The resulting polygon grid is an 'sfc' object, so it needs to be made {sf} in order for us to add the ID-attribute.

```{r message=FALSE, warning=FALSE}
grid_sf <- sf::st_make_grid(basins, n=c(10,10), what="polygons") %>%
    sf::st_sf()

grid_sf$ID <- 1:NROW(grid_sf)

plot(grid_sf)

```




### Intersection

Say we were interested in villages in a specific river basin. We can take an intersection of villages within that basin. First, let's load the villages from the file, and then inspect it together with the basins.

```{r message=FALSE, warning=FALSE}
# for CSC notebooks  replace ".." with "/home/rstudio/r-spatial-course"
villages <- sf::read_sf("../Data/villages.gpkg") %>%
    sf::st_transform(32648)

map <- mapview::mapview(villages) + basins
map

#select a basin and get all villages intersecting
( vil_sel <- dplyr::filter(basins, HYBAS_ID == 4061080120) %>%
    sf::st_intersection(villages) )

# If we only want HYBAS_ID from the basins data frame:
( vil_sel <- basins %>% 
    dplyr::select(HYBAS_ID) %>% 
    dplyr::filter(HYBAS_ID == 4061080120) %>%
    sf::st_intersection(villages) )

# we can use pipes inside functions as well
plot(dplyr::filter(basins, HYBAS_ID == 4061080120) %>% 
       dplyr::select(HYBAS_ID), 
     reset=FALSE)
plot(sf::st_geometry(vil_sel), pch=16, cex=1, col='white', add=TRUE)

```



### Clipping

Clipping can be done with st_intersection as well. However, for proper clipping we need to have a single polygon feature geometry. For instance, if we wish to cut the polygon grid to the basins dataset, we need to take union of basins and use it to intersect the grid.

```{r message=FALSE, warning=FALSE}
# Does not clip proper
grid_clip <- sf::st_intersection(grid_sf, basins)
plot(grid_clip, max.plot=2)

# Clips proper
grid_clip <-sf:: st_intersection(grid_sf, st_union(basins))
plot(grid_clip, max.plot=2)
```



## Exercises

1. Load Laos' admin areas, and *st_transform* admin to *"+proj=aea"* coordinate system (Albers Equal Area Projection). ggplot it.

```{r}

```


2. 

* *Select* NAME_1 attribute from admin areas, and MAIN_BAS from major basins. 
* Transform admin areas to EPSG:32648. 
* Take an intersection between basins and admin areas. 
* Add new attribute called area by using *st_area(geometry)* inside *mutate*.

```{r}

```

3.

* Take the intersection created in task 2. 
* Try to find out how many river basins intersect each province. Hint: use *group_by* and *summarise*.

```{r}

```

4. Find out what *st_area()* and *st_length()* do. Attach their results to a line or polygon features. You can use base R commands, *dplyr::mutate()*, or use the handy function *tibble::add_column()*.
```{r}

```

