---
title: 'CSC Spatial Analysis with R: Day 1, exercise answers'
author: "Marko Kallio"
date: "10 May 2023"
output:
  html_document:
    df_print: paged
    theme: flatly
---


# Exercises
## D1S1

Write out commands which

1.

* *read* in helsinki_buildings.gpkg from the data folder
* *select* a variable
* *filter* a subset of the buildings based on the variable you chose

```{r}
read_sf("../Data/helsinki_buildings.gpkg") %>% 
  dplyr::select(polyg_ala_m2) %>% 
  dplyr::filter(polyg_ala_m2 < 100)
```

2.

* *read* in helsinki_buildings.gpkg from the data folder
* *select* I_KERROSALA and D_ASHUOALA (total building sq m2, and sq m2 used for apartment space)
* *mutate* a new variable **proportion**, which is the proportion of D_ASHUOALA from the total I_KERROSALA.
* *mutate* a new variable *apartment_building* which has value 1 if **proportion** is higher than 0.5, and 0 if below. Hint: you can use **ifelse(condition, true, false)** inside *mutate*
* *group by* apartment_building and *summarise* the mean D_ASHUOALA for the two classes

```{r}
read_sf("../Data/helsinki_buildings.gpkg") %>% 
  dplyr::select(I_KERROSALA, D_ASHUOALA) %>% 
  dplyr::mutate(proportion = D_ASHUOALA / I_KERROSALA,
         apartment_building = ifelse(proportion >= 0.5, 1, 0)) %>%
  dplyr::group_by(apartment_building) %>% 
  st_set_geometry(NULL) %>% 
  summarise(mean(D_ASHUOALA))
  
# st_buffer(dist = 0)
# st_set_geometry(NULL)
# st_make_valid()

```

3.

* *read* in helsinki_buildings.gpkg from the data folder
* create a new attribute called *"ID"*, and give it values *1:nrow(buildings)*
* *select* ID from the buildings
* *filter* five (any five) buildings. You can use dplyr::filter(), giving it an argument of row numbers, e.g. 1:5
* create another set with another five buildings
* Attach the two sets together and plot it using plot() command

```{r}
buildings1 <- read_sf("../Data/helsinki_buildings.gpkg") %>% 
  dplyr::mutate(ID = 1:n()) %>% 
  dplyr::select(ID) %>% 
  slice(1:5) # alternative here would be slice()

buildings2 <- read_sf("../Data/helsinki_buildings.gpkg") %>% 
  dplyr::mutate(ID = 1:n()) %>% 
  dplyr::select(ID) %>% 
  tail(5)

rbind(buildings1, buildings2) %>% 
  plot
```

## D1S2

1. 

* *read* in helsinki_buildings.gpkg and Kaupunginosajako.kml from the data folder
* Create a ggplot which shows the districts and buildings.
* visualize districts as dashed line (*linetype = 2*) in the geom_sf() function call.
* give the districs a white or transparent *fill* colour.
* try different *linewidth*  and *color* for the district boundaries.

```{r}
buildings <- read_sf("../Data/helsinki_buildings.gpkg")
buildings

districts <- read_sf("../Data/Kaupunginosajako.kml")
districts

ggplot() +
  geom_sf(data = districts, fill = "transparent", linetype = 2,
          linewidth = 0.5, color = "grey50") +
  geom_sf(data = buildings)
```

2. Create another map zooming into one full district. Hint: Use *filter* to pick one district, and get the bounding box using function *st_bbox()*.

```{r}
?filter
library(dplyr)
bbox <- dplyr::filter(districts, nimi_fi == "MEILAHTI") %>% 
  st_bbox()
bbox

ggplot() +
  geom_sf(data = districts, fill = "transparent", linetype = 2,
          linewidth = 1, color = "red") +
  geom_sf(data = buildings) +
  coord_sf(xlim = c(bbox[1],bbox[3]), ylim = c(bbox[2],bbox[4]))
```

## D1S3

1. Load districts (Kaupunginosajako.kml), and *st_transform* districts to *"EPSG:3067"* coordinate system. ggplot it.

```{r}
districts <- read_sf("../Data/Kaupunginosajako.kml") %>% 
  st_transform(3067) 
districts

library(ggplot2)
ggplot() +
  geom_sf(data = districts)
```


2. 

* *Select* nimi_fi attribute from the districts 
* load the buildings data and select polyg_ala_m2 
* Take an intersection between districts and buildings. 
* Add new attribute called area by using *st_area(geometry)* (or *st_area(geom)*) inside *mutate*.

```{r}
districts <- select(districts, nimi_fi)
buildings <- read_sf("../Data/helsinki_buildings.gpkg") %>% 
  select(polyg_ala_m2) %>% 
  st_make_valid()

intersection <- st_intersection(districts, buildings) %>% 
  mutate(area = st_area(geometry))
intersection

```

3.

* Continuing using the intersection created in task 2.
* Try to find out how many buildings intersect each district. Hint: use *group_by* and *summarise*.

```{r}
intersection %>% 
  group_by(nimi_fi) %>% 
  summarise(n())

table(intersection$nimi_fi)
```



4. Find out what *st_area()* and *st_length()* do. Attach their results to a polygon or line (e.g. by casting districts from polygon to linestring). You can use base R commands, *dplyr::mutate()*, or use the handy function *tibble::add_column()*.
```{r}
dist_areas <- st_area(districts)
dist_areas
districts <- tibble::add_column(districts, area_m2 = dist_areas)
districts
districts$area_m2_2 <- dist_areas
districts
```

## D1S4

1. Write an R script which

* loads the building dataset
* transforms CRS to EPSG:3067
* selects one (any) building
* takes a buffer of 1 km
* counts how many buildings fall within the buffer
* plots the buffer and the buildings

```{r}
library(sf)

# Some buildings seem to have invalid (self-intersecting) geometry, so therefore here an extra step here for fixing that.
buildings <- read_sf("../Data/helsinki_buildings.gpkg") %>% 
  st_transform(3067) %>% 
  st_make_valid()
  
# Take random 
one_building <- buildings[450,]

one_building_1km_buffer <- one_building %>%
    sf::st_buffer(dist = 1000)

buildings_selected <- sf::st_intersection(one_building_1km_buffer, buildings)
nrow(buildings_selected)

plot(one_building_1km_buffer[,"valmvuosi"], reset=FALSE)
plot(builidings_selected[,"valmvuosi"], col='white', add=TRUE)


```

2. Do the same as above, but take a random sample of 10 buildings.

* use ggplot and use colour aestethic to visualize the number of villages inside the buffers.

```{r}
library(ggplot2)

ten_buildings <- buildings %>% #replace buildings here with the name of your variable from task 1.
  sample_n(10) # take random sample

ten_buildings_1km_buffer <- ten_buildings %>%
    sf::st_buffer(dist = 1000)

ten_buildings_selected <- sf::st_intersection(ten_buildings_1km_buffer, buildings)

builiding_counts <- sf::st_intersects(ten_buildings_1km_buffer, buildings, sparse=FALSE)
builiding_counts <- apply(builiding_counts, MARGIN = 1, sum)
ten_buildings_1km_buffer$builiding_counts <- builiding_counts

ten_buildings_1km_buffer

ggplot(ten_buildings_1km_buffer) + 
  geom_sf(aes(fill = builiding_counts)) +
  geom_sf(data = ten_buildings_selected) 

```