---
title: "CSC Spatial Analysis with R: Day 2 Session 3 - Spatial models"
author: "Marko Kallio"
date: "11 May 2023"
output:
  html_document:
    df_print: paged
    theme: flatly
---


## Spatial Autoregressive (SAR) Models

Spatial autocorrelation may be included in regression models using Spatial Autoregressive Models (among other options). SAR models can be defined as

**y** = p**Wy** + **Xb** + e 

where **y** is dependent variable, p is the strength of spatial association (autocorrelation), **W** is a spatial weights matrix, **X** is the matrix of variables, **b** is a vector of regression coefficients and e is the error.

```{r}
# library(spdep)
library(spatialreg)
library(sf)
library(dplyr)

set.seed(20210102) # for reproducibility

villages <- sf::read_sf("../Data/villages.gpkg") %>% 
    sf::st_transform(32648) %>% 
    dplyr::sample_n(500)

admin <- sf::read_sf("../Data/LAO_adm1.shp") %>% 
  sf::st_transform(32648)

vil_voronoi <- villages %>% 
      sf::st_union() %>% 
      sf::st_voronoi() %>% 
      sf::st_cast() %>% 
      sf::st_sf() %>% 
      sf::st_join(villages) %>% 
      sf::st_intersection(st_union(admin))


# create 'listw' (weight list) object from point distance matrix  
dmat <- 1/sf::st_distance(villages) %>% 
  unclass() 
diag(dmat) <- 0 # set matrix diagonal as 0
dmat_listw <- spdep::mat2listw(dmat, style = "W")

vil_data <- villages %>% sf::st_set_geometry(NULL)
```


```{r}
sar <- spatialreg::lagsarlm(PovertyRate ~  LiteracyRate + HumanFootprint, 
         data=vil_data, 
         dmat_listw)
summary(sar)

cor(villages$PovertyRate, sar$fitted.values)^2

# compare to OLS
linmod <- lm(PovertyRate ~  LiteracyRate + HumanFootprint, 
         data=vil_data)
summary(linmod)

cor(villages$PovertyRate, linmod$fitted.values)^2
```





## Geographically weighted regression

Geographically weighted regression (GWR) is a method which allows accounting for spatial non-heterogeneity in regression analysis. In practise, it performs a regular regression on a weighted sample of observations. It is closely related to locally weighted regression (LWR) with a difference that LWR weights observations in attribute space (which can include spatial information), while GWR weights observations in geographical space, and attribute information does not play a role in the weights calculation. 

Weighting of observations in GWR is based on some kernel method, based on bandwidth (the extent of the spatial phenomenon). The bandwidth can be computationally optimised, or be based on expert knowledge.

In R, GWR has been implemented at least in {GWmodel}, {spgwr}, and in the upcoming {gwverse}, suite of packages. Here we use {GWmodel}. It does not support 'sf' objects, so we need to use 'sp' classes. 

```{r message=FALSE, warning=FALSE}
library(GWmodel, quietly=TRUE)
#transform
villages_sp <- as(villages, "Spatial")

# calculate distance matrix and compute optimal bandwidth for GWR
dmat <- gw.dist(dp.locat=coordinates(villages_sp))
bw <- bw.gwr(PovertyRate ~  LiteracyRate + HumanFootprint,  
             villages_sp, kernel = "gaussian")
```

The optimal cross-validated bandwidth for this model seems to be approximately 36.5 km. We can then perform GWR using this bandwidth:

```{r message=FALSE, warning=FALSE}
vil_gwr <- gwr.basic(PovertyRate ~  LiteracyRate + HumanFootprint, 
                     villages_sp, kernel = "gaussian", bw = bw, dMat = dmat)

vil_voronoi$pov_GWR <- vil_gwr$SDF$y
plot(vil_voronoi[,"pov_GWR"])
plot(vil_voronoi[,"PovertyRate"])


```

The spatial patterns seems to be off somewhat.

```{r message=FALSE, warning=FALSE}
vil_gwr

```


Resulting object from GWR contains a lot of useful information:

* GW.arguments - a list class object including the model fitting parameters for generating the report file
* GW.diagnostic	- a list class object including the diagnostic information of the model fitting
* lm - an object of class inheriting from “lm”, see lm.
* SDF - a SpatialPointsDataFrame (may be gridded) or SpatialPolygonsDataFrame object (see package “sp”) integrated with fit.points,GWR coefficient estimates, y value,predicted values, coefficient standard errors and t-values in its "data" slot.
* timings	- starting and ending time.
* this.call	- the function call used.
* Ftest.res - results of Leung's F tests when F123.test is TRUE.




## Exercise

1. 

* read in villages, and transform it to EPSG:32648
* take a sample of 500 random villages
* cast it as "Spatial"
* calculate a distance matrix
* optimise bandwidth for GWR for model **PovertyRate ~ 4 variables of your choosing**
* run GWR with the given model
* plot the residuals (can be obtained from *vil_gwr$SDF$residual*)

```{r}

```


2.

* Estimate the same model using a SAR model

```{r}

```

