---
title: "CSC Spatial Analysis with R: D2S1 - Spatial autocorrelation"
author: "Marko Kallio"
date: "11 May 2023"
output:
  html_document:
    df_print: paged
    theme: flatly
---


Let's start the second day making sure we have everything required installed.

```{r}
inst_packages <- installed.packages()
for (package in c("dplyr", "sf", "spdep", "ncf", "units", "inlmisc",
                  "spatstat", "fpc", "GWmodel", "NbClust", "patchwork", 
                  "spatialreg")) {
    test <- package %in% inst_packages[,1]
    if (!test) {
        install.packages(package)
    }
}
rm(package, inst_packages, test)
```


# Spatial autocorrelation

Spatial autocorrelation occurs when points, or their attributes are not randomly distributed in space, but exhibit some kind of spatial relationship. Spatial autocorrelation can be positive (the attribute values are clustered), negative (values are dispersed), or there is no spatial autocorrelation (value distribution is spatially random). 



## Moran's Index
The most common metric to measure spatial autocorrelation is Moran's I. In R, we can use package 'spdep' to compute Moran's I. First, we must define a weight matrix. The Index takes values between -1 and 1, with 0 meaning no spatial autocorrelation. Commonly any value higher 0.30 or lower than -0.30 is interpreted as "strong" spatial autocorrelation. The weight matrix can affect the Index value quite a bit, so we examine here three possible ways (among many others) to construct it: distance matrix, k nearest neighbours, and neighborhoods. 

```{r message=FALSE, warning=FALSE}
library(sf, quietly=TRUE)
library(dplyr, quietly=TRUE)

set.seed(20230511)
villages <- read_sf("../Data/villages.gpkg") %>% 
    st_transform(32648) %>% 
    sample_n(250)

villages
plot(villages)

library(spdep)

# first create 'listw' (weight list) object from point distance matrix  
dmat <- 1/sf::st_distance(villages)^2 %>% 
  unclass() #unclassing because st_distance returns a 'unit' class object. Operations with 'unit' class require both operands to be of class 'unit'.
class(dmat)
dmat
diag(dmat) <- 0 # set matrix diagonal as 0
dmat

dmat_listw <- spdep::mat2listw(dmat, style = "W")
dmat_listw$neighbours[[1]]
sum(dmat_listw$weights[[1]])

( moran <- spdep::moran.test(villages$LiteracyRate, dmat_listw) )
```

```{r}
# weights based on k nearest neighbours
knn_listw <- sf::st_coordinates(villages) %>% 
    spdep::knearneigh(k=10) %>% 
    spdep::knn2nb() %>% # create neighbour list
    spdep::nb2listw(style = "W") # create weights list from neighbour object
knn_listw$neighbours[[1]]
knn_listw$weights[[1]]
spdep::moran.test(villages$PovertyRate, knn_listw)

```

```{r}
# weights based on polygon neighborhoods
admin <- sf::read_sf("../Data/LAO_adm1.shp") %>%
    sf::st_transform(st_crs(villages))

vil_voronoi <- villages %>% 
      sf::st_union() %>% 
      sf::st_voronoi() %>% 
      sf::st_cast() %>% 
      sf::st_sf() %>% 
      sf::st_join(villages) %>% 
      sf::st_intersection(st_union(admin))
plot(vil_voronoi)


nb <- spdep::poly2nb(vil_voronoi)
summary(nb)

# plot(st_geometry(vil_voronoi), col='grey' )
# plot(nb, st_coordinates(villages), add=TRUE)

nb_listw <- spdep::nb2listw(nb, style = "W")

spdep::moran.test(villages$PovertyRate, nb_listw)

plot(vil_voronoi[,"PovertyRate"])


```

## Local Moran's I

Local Moran's I is also simple to compute using the 'spdep' package. It is important to note that local measure of Moran's I does not necessarily lie within -1 and 1. Read more: Luc Anselin's 1995 paper "Local Indicators of Spatial Association—LISA" seems to be open access. http://onlinelibrary.wiley.com/doi/10.1111/j.1538-4632.1995.tb00338.x/full. 

```{r message=FALSE, warning=FALSE}
local_moran <- spdep::localmoran(villages$PovertyRate, 
                                 nb_listw, 
                                 alternative = "two.sided") %>% 
    dplyr::as_tibble()

# Local Moran's Index
local_moran

# plot
vil_voronoi <- vil_voronoi %>% 
  dplyr::mutate(Morans_I = as.vector(local_moran$Ii), 
                Morans_p = as.vector(local_moran$`Pr(z != E(Ii))`)) # # p-value of local moran statistic. Notice that the column name has spaces, which means we need to refer to it with backticks. In finnish language keyboard, press shift + the button to the left of backspace.
plot(vil_voronoi[,"Morans_I"])
plot(vil_voronoi[,"Morans_p"])


```

```{r}
library(ggplot2)
p1 <- ggplot() + 
  geom_sf(data = vil_voronoi, aes(fill = Morans_I)) +
  scale_fill_gradient2(low="red", mid = 'white', high = 'blue') +
  labs(title = "Moran's I")

p2 <- ggplot() + 
  geom_sf(data = vil_voronoi, aes(fill = Morans_p)) +
  scale_fill_gradient(low="red",high = 'blue') +
  labs(title = "Moran's I, p-value")

library(patchwork)
p1 + p2
```



## Correlogram


We use the package ncf here to compute the correlogram. A correlogram estimates spatial autocorrelation for different distance bands, rather than trying to estimate a single Moran's I for the entire dataset. 
```{r point correlogram}
library(ncf)

coordinates <- sf::st_coordinates(villages)
coordinates

corgram <- ncf::correlog(coordinates[,1], coordinates[,2], # point coordinates
                         villages$PovertyRate, # value for estimation
                    increment=20000, # distance lags, here 100 km
                    resamp=50, # permutations for significance testing
                    na.rm=TRUE)
plot(corgram) # closed dots have p < 0.05

( data <- data.frame(distance = corgram$mean.of.class,
                       moran = corgram$correlation,
                       p = corgram$p) )
```





## Exercises 

1. 

* read villages.gpkg and transform it into *EPSG:32648*
* *filter* province Savannakhet in villages dataset. 
* compute a *1/st_distance* matrix for the subset of villages, and create a *listw* object from it
* compute Moran's I for an attribute of your own choice
* draw a ggplot

```{r}
villages <- read_sf("../Data/villages.gpkg") %>% 
  st_transform(32648) %>% 
  filter(prov_name == "Savannakhet")

dmat <- 1/sf::st_distance(villages)^2 %>% 
  unclass() #unclassing because st_distance returns a 'unit' class object. Operations with 'unit' class require both operands to be of class 'unit'.

diag(dmat) <- 0 # set matrix diagonal as 0
dmat_listw <- spdep::mat2listw(dmat, style = "W")

( moran <- spdep::moran.test(villages$PovertyRate, dmat_listw) )

ggplot() +
  geom_sf(data = villages, aes(color = PovertyRate))

```


2.

* Take subset of villages from exercise 1
* Compute *listw* object from *k*-nearest neighbours (choose *k* yourself)
* compute local moran using the subset and knn-listw for Savannakhet.

```{r}
knn_listw <- sf::st_coordinates(villages) %>% 
    spdep::knearneigh(k=10) %>% 
    spdep::knn2nb() %>% # create neighbour list
    spdep::nb2listw(style = "W") # create weights list from neighbour object

local_moran <- spdep::localmoran(villages$PovertyRate, 
                                 knn_listw, 
                                 alternative = "two.sided") %>% 
    dplyr::as_tibble()

# Local Moran's Index
local_moran

# plot
villages <- villages %>% 
  dplyr::mutate(Morans_I = as.vector(local_moran$Ii), 
                Morans_p = as.vector(local_moran$`Pr(z != E(Ii))`)) # # p-value of local moran statistic. Notice that the column name has spaces, which means we need to refer to it with backticks. In finnish language keyboard, press shift + the button to the left of backspace.

p1 <- ggplot() + 
  geom_sf(data = villages, aes(color = Morans_I)) +
  scale_fill_gradient2(low="red", mid = 'white', high = 'blue') +
  labs(title = "Moran's I")

p2 <- ggplot() + 
  geom_sf(data = villages, aes(color = Morans_p)) +
  scale_fill_gradient(low="red",high = 'blue') +
  labs(title = "Moran's I, p-value")

library(patchwork)
p1 + p2
```


3. 

* compute a correlogram for Savannakhet for a variable of your choice.

```{r}

coordinates <- sf::st_coordinates(villages)
coordinates

corgram <- ncf::correlog(coordinates[,1], coordinates[,2], # point coordinates
                         villages$PovertyRate, # value for estimation
                    increment=20000, # distance lags, here 100 km
                    resamp=10, # permutations for significance testing
                    na.rm=TRUE)
plot(corgram) # closed dots have p < 0.05

( data <- data.frame(distance = corgram$mean.of.class,
                       moran = corgram$correlation,
                       p = corgram$p) )
```







# EXTRA TOPICS

# Point patterns and densities

Point patterns can be extensively analysed using the 'spatstat' package. Here are some examples of point pattern-related functions from the package.


## Quadrats

Quadrat analysis can be used to sample point patterns in spatial sampling areas to test for spatial distribution pattersn, randomness and clustering. In practise, we already did this in the end of previous lesson with intersections of village points to the polygon grid.
The functions in spatstat require the point pattern to be of class 'ppp', so we create that first before running the quadratcount() function.

```{r message=FALSE, warning=FALSE}
library(spatstat, quietly = TRUE)

# create village points of class 'ppp' (point pattern) from 'spatstat' package
bbox <- st_bbox(admin)
vil_coords <- st_coordinates(villages)

window <- owin(c(bbox["xmin"], bbox["xmax"]), c(bbox["ymin"], bbox["ymax"]))
vil_ppp <- ppp(vil_coords[,1], vil_coords[,2], window = window)

quadrats <- quadratcount(vil_ppp, nx=10, ny=10)

plot(vil_ppp, pch=20, col="grey")
plot(quadrats, add=TRUE)

```


The function intensity() can then be run to get the quadrat intensity, which divides the quadrat counts by the tile areas. This yields the average density of points per unit area in each tile.
```{r message=FALSE, warning=FALSE}
Q_density <- intensity(quadrats, image=TRUE)
plot(Q_density) # number of points per m2
```



## Kernel density estimation

Kernel density estimation can be used for instance to evaluate point clustering. It outputs a continuous surface based on point count in the neighborhoods. In 'spatstat', this can be computed with density().
```{r message=FALSE, warning=FALSE}
# 10 km density
K_dens <- density(vil_ppp, sigma=10000)
plot(K_dens)
plot(admin, col='grey80', add=TRUE)
contour(K_dens, add=TRUE) # add contours to the plot
```





## Other point pattern functions

There are many more functions available for extended point pattern analysis, such as nearest neighbours, K-, L-, F-, or G-functions. These are all included in the spatstat package. 

```{r}
# Average nearest neighbour distance
nn <- nndist(vil_ppp, k=1)
mean(nn)

# K-, L-, F-, or G-functions
vil_G <- Gest(vil_ppp)
vil_L <- Lest(vil_ppp)
vil_K <- Kest(vil_ppp)
vil_F <- Fest(vil_ppp)
plot(vil_G)
```
