---
title: 'CSC Spatial Analysis with R: Day 1, exercise answers'
author: "Marko Kallio"
date: "10 May 2023"
output:
  html_document:
    df_print: paged
    theme: flatly
---

# Exercises
## D2S1

1. 

* read villages.gpkg and transform it into *EPSG:32648*
* *filter* province Savannakhet in villages dataset. 
* compute a *1/st_distance* matrix for the subset of villages, and create a *listw* object from it
* compute Moran's I for an attribute of your own choice
* draw a ggplot

```{r}
villages <- read_sf("../Data/villages.gpkg") %>% 
  st_transform(32648) %>% 
  filter(prov_name == "Savannakhet")

dmat <- 1/sf::st_distance(villages)^2 %>% 
  unclass() #unclassing because st_distance returns a 'unit' class object. Operations with 'unit' class require both operands to be of class 'unit'.

diag(dmat) <- 0 # set matrix diagonal as 0
dmat_listw <- spdep::mat2listw(dmat, style = "W")

( moran <- spdep::moran.test(villages$PovertyRate, dmat_listw) )

ggplot() +
  geom_sf(data = villages, aes(color = PovertyRate))

```


2.

* Take subset of villages from exercise 1
* Compute *listw* object from *k*-nearest neighbours (choose *k* yourself)
* compute local moran using the subset and knn-listw for Savannakhet.

```{r}
knn_listw <- sf::st_coordinates(villages) %>% 
    spdep::knearneigh(k=10) %>% 
    spdep::knn2nb() %>% # create neighbour list
    spdep::nb2listw(style = "W") # create weights list from neighbour object

local_moran <- spdep::localmoran(villages$PovertyRate, 
                                 knn_listw, 
                                 alternative = "two.sided") %>% 
    dplyr::as_tibble()

# Local Moran's Index
local_moran

# plot
villages <- villages %>% 
  dplyr::mutate(Morans_I = as.vector(local_moran$Ii), 
                Morans_p = as.vector(local_moran$`Pr(z != E(Ii))`)) # # p-value of local moran statistic. Notice that the column name has spaces, which means we need to refer to it with backticks. In finnish language keyboard, press shift + the button to the left of backspace.

p1 <- ggplot() + 
  geom_sf(data = villages, aes(color = Morans_I)) +
  scale_fill_gradient2(low="red", mid = 'white', high = 'blue') +
  labs(title = "Moran's I")

p2 <- ggplot() + 
  geom_sf(data = villages, aes(color = Morans_p)) +
  scale_fill_gradient(low="red",high = 'blue') +
  labs(title = "Moran's I, p-value")

library(patchwork)
p1 + p2
```


3. 

* compute a correlogram for Savannakhet for a variable of your choice.

```{r}

coordinates <- sf::st_coordinates(villages)
coordinates

corgram <- ncf::correlog(coordinates[,1], coordinates[,2], # point coordinates
                         villages$PovertyRate, # value for estimation
                    increment=20000, # distance lags, here 100 km
                    resamp=10, # permutations for significance testing
                    na.rm=TRUE)
plot(corgram) # closed dots have p < 0.05

( data <- data.frame(distance = corgram$mean.of.class,
                       moran = corgram$correlation,
                       p = corgram$p) )
```


## D2S2


1. *filter* villages with prov_name **Savannakhet**
2. create **cluster_data** object with *select* 3 variables
3. *st_set_geometry* to NULL
4. Apply k-means clustering with 4 clusters
5. Attach the clusters to Savannakhet villages (or voronoi, if you create them)
6. ggplot the clusters. Optionally a boxplot too.

```{r}
savannakhet <- read_sf("../Data/villages.gpkg") %>% 
  filter(prov_name == "Savannakhet")

cluster_data <- savannakhet %>% 
  select(HumanFootprint, SoilDegradation, IncomeDependency) %>% 
  st_set_geometry(NULL)

clusters <- kmeans(cluster_data, 4)

savannakhet$cluster <- factor(clusters$cluster)

ggplot(savannakhet) + 
  geom_sf(aes(color = cluster)) +
  geom_sf(data = admin %>% filter(NAME_1 == "Savakhannet"), fill = "transparent", colour = "black") +
  scale_fill_brewer(palette = "Accent") +
  labs(title = "Cluster") +
theme_bw()
admin
```


7. Do the same as above, but using hierarchical clustering and compute the distance matrix with the *scaled_dist()* function we defined earlier.

```{r}
dmat <- scaled_dist(savannakhet, scale_range = c(1,3))


# ------------------------------------------------------------------------------
# hierarchical clustering with scaled distance metric
clusters <- hclust(dmat, method = "complete")
plot(clusters)
clusters_cut <- cutree(clusters, k=4)

# number of points in each cluster
print(table(clusters_cut))


savannakhet$cluster_scaled <- factor(clusters_cut)

ggplot(savannakhet) + 
  geom_sf(aes(color = cluster_scaled)) +
  geom_sf(data = admin %>% filter(NAME_1 == "Savakhannet"), fill = "transparent", colour = "black") +
  scale_fill_brewer(palette = "Accent") +
  labs(title = "Cluster_scaled")
```


## D2S3

1. 

* read in villages, and transform it to EPSG:32648
* take a sample of 500 random villages
* cast it as "Spatial"
* calculate a distance matrix
* optimise bandwidth for GWR for model **PovertyRate ~ 4 variables of your choosing**
* run GWR with the given model
* plot the residuals (can be obtained from *vil_gwr$SDF$residual*)

```{r}
villages <- read_sf("../Data/villages.gpkg") %>% 
  st_transform(32648) %>% 
  sample_n(500) 
vil_sp <- as(villages, "Spatial")

dmat <- gw.dist(dp.locat=coordinates(vil_sp))
names(villages)
bw <- bw.gwr(PovertyRate ~  wIrrigatedArea + SoilDegradation + IncomeDependency + DrinkWaterSource,   
             vil_sp, kernel = "gaussian")

vil_gwr <- gwr.basic(PovertyRate ~ wIrrigatedArea + SoilDegradation + IncomeDependency + DrinkWaterSource,   
                     vil_sp, kernel = "gaussian", bw = bw, dMat = dmat)
vil_gwr

gwr_sf <- vil_gwr$SDF %>% st_as_sf()

plot(gwr_sf[,"residual"])
```


2.

* Estimate the same model using a SAR model

```{r}
# create 'listw' (weight list) object from point distance matrix  
dmat <- 1/sf::st_distance(villages)^2 %>% 
  unclass() 
diag(dmat) <- 0 # set matrix diagonal as 0
dmat_listw <- spdep::mat2listw(dmat, style = "W")

vil_data <- villages %>% sf::st_set_geometry(NULL)

sar <- spatialreg::lagsarlm(PovertyRate ~ wIrrigatedArea + SoilDegradation + IncomeDependency + DrinkWaterSource,    
         data=vil_data, 
         dmat_listw)
summary(sar)

spdep::moran.test(gwr_sf$residual, dmat_listw)

```